Neural Connectome of the Ctenophore Statocyst

### 2025-05-22 manuscript/Manuscript.qmd
MS: discussion edits

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve net (ANNs)—with multiple nuclei and a morphology distinct from thesubepithelial nerve net (SNN) with a “beads-on-a-string” organization reported in one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.

Remarkably, ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells. Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system. These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity [@keijzer2013; @dewiljes2015; @Veraszto2017].

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014]. However, the combination of circuit architecture and behavioural outputs provides the first functional insight into the role of neurons in a ctenophore.

>>>>

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron---here referred to as aboral nerve net (ANNs)---with multiple nuclei. The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion. The presence of two SNN-type neurons more laterally in the aboral organ in our dataset (Figure 2--figure supplement1) further underline the distinct identity of ANNs.

ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells. Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system. These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].

What could be the function of the aboral nerve net? We found that the anatomical differences in the innervation patterns of ANNs across the sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This suggests that individual ANNs may play opposing or complementary roles in regulating ciliary beating---potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014]. However, the combination of circuit architecture and behavioural outputs provides the first functional insight into the role of neurons in a ctenophore.
